---
title: 'Reproducible Research: Peer Assessment 2'
author: "Gaoqi Liang"
date: "26 September 2015"
output: pdf_document
---
# Impact of Severe Weather Events on Public Health and Economy in the United States
## Synopsis
This document is used to analyze the impact of stroms and other severe weather events on public health and economy.The data analysis mainly address the following questions: (1) Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health. (2) Across the United States, which types of events have the greatest economic consequences.

## Data Processing
```{r, echo=FALSE}
library(plyr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(R.utils)
require(gridExtra)

```

Download data, unzip data, and read data
```{r, echo=TRUE}
## download files
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
file <- "stormdata.csv.bz2"
if(!file.exists(file)){
  download.file(url, file)
}

## unzip and create folders (if those ain't exist)
unzip <- bzfile("stormData.csv.bz2", "r")
stormData <- read.csv(unzip, stringsAsFactors = FALSE)
close(unzip)

## get familiar with data
dim(stormData)
names(stormData)
head(stormData)

```

List the stormData by year from 1950 to 2010
```{r, echo=TRUE}
if (dim(stormData)[2] == 37) {
  stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(stormData$year, breaks = 50)
```

We select the data after 1980. Because "In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete." 
```{r, echo=TRUE}
## subset the data from 1980 to 2010
storm <- stormData[stormData$year > 1980, ]
dim(storm)
```


## Impact of Severe Weather Events on Public Health
Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r, echo=TRUE}
## sum of fatalities by event type, sort them 
fatalities <- aggregate(FATALITIES~EVTYPE,storm,sum)
sort.fatalities <- fatalities[order(-fatalities$FATALITIES),]
fatalities30 <- sort.fatalities[1:30,]
```

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r, echo=TRUE}
## sum of injuries by event type, sort them 
injuries <- aggregate(INJURIES~EVTYPE,storm,sum)
sort.injuries <- injuries[order(-injuries$INJURIES),]
injuries30<- sort.injuries[1:30,]

head(fatalities30)
head(injuries30)

```

## Impact of Severe Weather Events on Economy
Across the United States, which types of events have the greatest economic consequences?
We focus on the property damage and crop damage that types of events cause in order to see the most costly.
Sum of property damage multiplied by their respective exponent by event type, sort them
```{r, echo=TRUE}
unique(storm$PROPDMGEXP)

##lets make them numeric by the multipliers in data documentation
storm$PROPEXPN[storm$PROPDMGEXP ==""] <- 1
storm$PROPEXPN[storm$PROPDMGEXP =="-"] <- 0
storm$PROPEXPN[storm$PROPDMGEXP =="?"] <- 0
storm$PROPEXPN[storm$PROPDMGEXP =="+"] <- 0
storm$PROPEXPN[storm$PROPDMGEXP =="0"] <- 1
storm$PROPEXPN[storm$PROPDMGEXP =="1"] <- 1
storm$PROPEXPN[storm$PROPDMGEXP =="2"] <- 100
storm$PROPEXPN[storm$PROPDMGEXP =="3"] <- 1000
storm$PROPEXPN[storm$PROPDMGEXP =="4"] <- 10000
storm$PROPEXPN[storm$PROPDMGEXP =="5"] <- 100000
storm$PROPEXPN[storm$PROPDMGEXP =="6"] <- 1e+06
storm$PROPEXPN[storm$PROPDMGEXP =="7"] <- 1e+07
storm$PROPEXPN[storm$PROPDMGEXP =="8"] <- 1e+08
storm$PROPEXPN[storm$PROPDMGEXP =="B"] <- 1e+09
storm$PROPEXPN[storm$PROPDMGEXP =="h"] <- 100
storm$PROPEXPN[storm$PROPDMGEXP =="H"] <- 100
storm$PROPEXPN[storm$PROPDMGEXP =="K"] <- 1000
storm$PROPEXPN[storm$PROPDMGEXP =="m"] <- 1e+06
storm$PROPEXPN[storm$PROPDMGEXP =="M"] <- 1e+06

storm$PROPDMGWITHEXP <- storm$PROPDMG * storm$PROPEXPN

pdmg <- aggregate(PROPDMGWITHEXP~EVTYPE,storm,sum)
sort.pdmg <- pdmg[order(-pdmg$PROPDMGWITHEXP),]
pdmg30 <- sort.pdmg[1:30,]

```


Sum of crop damage multiplied by their respective exponent by event type, sort them 
```{r, echo=TRUE}
unique(storm$CROPDMGEXP) ##to see the levels

## lets make them numeric by the multipliers in data documentation
storm$CROPEXPN[storm$CROPDMGEXP == ""] <- 1
storm$CROPEXPN[storm$CROPDMGEXP == "?"] <- 0
storm$CROPEXPN[storm$CROPDMGEXP == "0"] <- 1
storm$CROPEXPN[storm$CROPDMGEXP == "2"] <- 10
storm$CROPEXPN[storm$CROPDMGEXP == "B"] <- 1e+9
storm$CROPEXPN[storm$CROPDMGEXP == "k"] <- 1000
storm$CROPEXPN[storm$CROPDMGEXP == "K"] <- 1000
storm$CROPEXPN[storm$CROPDMGEXP == "m"] <- 1e+06
storm$CROPEXPN[storm$CROPDMGEXP == "M"] <- 1e+06

storm$CROPDMGWITHEXP <- storm$CROPDMG * storm$CROPEXPN

cdmg <- aggregate(CROPDMGWITHEXP~EVTYPE,storm,sum)
sort.cdmg <- cdmg[order(-cdmg$CROPDMGWITHEXP),]
cdmg30 <- sort.cdmg[1:30,]

```

## Results
```{r, echo=TRUE}
## List of the events with the largest number of fatalities.
fatalities30
## List of the events with the largest number of injuries.
injuries30

```

Graphs of the fatalities and injuries by event.
```{r, echo=TRUE}
library(ggplot2)
fatalitiesPlot <- ggplot(fatalities30,aes(EVTYPE,FATALITIES))+
                          geom_bar(stat="identity",fill="purple")+
                          coord_flip()+xlab("Weather Event Type")+
                          ylab("Fatalities")+
                          theme(axis.text.x = element_text(angle = 90, hjust = 1))
injuriesPlot <- ggplot(injuries30,aes(EVTYPE,INJURIES))+
                        geom_bar(stat="identity",fill="red")+
                        coord_flip()+
                        xlab("Weather Event Type")+
                        ylab("Injuries")+
                        theme(axis.text.x = element_text(angle = 90, hjust = 1))

## putting the plots in a single row 
library(gridExtra)

## Loading required package: grid (some problem)
grid.arrange(fatalitiesPlot,injuriesPlot,ncol = 2)
```

It is shown from the figure that tornadoes and excessive heat are the top 2 most significant impact on the public health in terms of injuries and fatalities. 


```{r, echo=TRUE}
# List of the events with the highest economical impact(considering properties).
pdmg30
# List of the events with the highest economical impact(considering crops).
cdmg30

```

Graphs of the property and crop damages by event.
```{r, echo=TRUE}
propertyPlot <- ggplot(pdmg30,aes(EVTYPE,PROPDMGWITHEXP))+
                geom_bar(stat="identity",fill="purple")+
                coord_flip()+
                xlab("Weather Event Type")+
                ylab("Property Damage")+
                theme(axis.text.x = element_text(angle = 90, hjust = 1))
cropPlot<- ggplot(cdmg30,aes(EVTYPE,CROPDMGWITHEXP))+
                geom_bar(stat="identity",fill="red")+
                coord_flip()+xlab("Weather Event Type")+
                ylab("Crop Damage")+
                theme(axis.text.x = element_text(angle = 90, hjust = 1))

##putting the plots in a single row 
library(gridExtra)

##putting the plots in a single row (some problem)
grid.arrange(propertyPlot,cropPlot,ncol = 2)

```
It is shown from the figure that floods are the leading cause of property damage followed by hurricanes/typhoons and storm surges. Leading cause of damage to crops are draughts and floods.

## Conclusion
It is shown from the data that (1) tornadoes and excessive heat are the top 2 most harmful weather events for public health. (2) Floods followed by hurricanes/typhoons and storm surges are the costliest severe weather events considering property damage. (3) Draughts and floods are the costliest considering crop damage.



